One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-08-02 doc/coverletter.tex
spell check

Enclosed you will find a manuscript entitled: ``A simple null model for inferences from network enrichment analysis''. The manuscript describes an in our minds natural null model for network enrichment analysis, as well as two implementations of the null model as an exact test and a sampling method.

Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.  In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters. We also demonstrate our implementations to render far better calibration than the state-of the-art method, BinoX.

@@

The manuscript is acompanied by an open github acount with all software and all tests that we conducted, available for reviewers, or the general public. The manuscript is an original contribution and none of the work it describes has been published elsewhere except for in abstract form. The manuscript will remain with you and will not be submitted elsewhere, until you made a decision as to its suitability for publication in PLOS ONE.

>>>>

Enclosed you will find a manuscript entitled: ``A simple null model for inferences from network enrichment analysis''. The manuscript describes an in our minds natural null model for network enrichment analysis (NEA), as well as two implementations of the null model as an exact test and a sampling method.

Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.  In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters. We also demonstrate our implementations to render far better calibration than the state-of-the-art method, BinoX.

@@

While the manuscript does not present a user-friendly software, it instead focuses on how to accurately assess error probabilities for NEA. We note that several methods in the field do not report calibrated statistics and hence are hard to compare. In our minds, accurate statistics is the first step in the process of improving the field of NEA.

The manuscript is accompanied by an open Github account with all software and all tests that we conducted, available for reviewers, or the general public. The manuscript is an original contribution and none of the work it describes has been published elsewhere except for in abstract form. The manuscript will remain with you and will not be submitted elsewhere until you made a decision as to its suitability for publication in PLOS ONE.

### 2018-08-02 doc/genesetdp.tex
change figure label
fix label
spell check

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*lukas.kall@scilifelab.se
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks. While the termonolgy vary in the litterature, we will here refer to such methods as Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations. Here we instead have designed one dynamic programming  algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We also implemented a random sampling method, carrying out the same task. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.

@@

A limiting factor of ORA is the quality of the databases defining pathways. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the termonolgy vary in the litterature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).

@@

Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}. We argue that such test are much easier implemented as tests of the query set, i.e. is the query set a randomly selected set of genes or is there an overrepresentation of well-connected genes in the query set?

@@

In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ (i.e. $\mathcal{Q} \subseteq \mathcal{G}$), and how they relate to a pathway, $\mathcal{P}=\{p_1 \ldots p_P\}$. In NEA the pathway maps to the genome through a network, ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise. The pathways are scored by summing up all connections, i.e. the pathway score can be expressed as,

@@

We wanted to evaluate our score under the null hypothesis, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''. This  formulation translates to calculating the distribution of scores that would occur if the $Q$ query genes are randomly selected from the genome, and see how frequently a score as extreme or more extreme than the current score would appear in that distribution. Hence, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$, which we will give two methods for below. Given such a score distribution, we can express a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score, $s$, as,

@@

    for a in range(len(k)):
        for s in range(max_s,-1,-1):
            for c in range(Q,-1,-1):
                for b in range(k[a],0,-1): # Stop at b=1
                    if c-b>=0 and s-a*b>=0:
                        N[s,c] += comb(k[a], b) * N[s-a*b,c-b]
    return N[:,q]
\end{lstlisting}
There is a memory efficient implementation. We first note that

@@

To test the consistency of the methods, we also investigated the correspondance between the $p$~values reported by GeneSetDP and GeneSetMC. As can be seen in Figure~\ref{fig:pscatter}, the results are consistent as expected.

\begin{figure}[htb]
  \begin{center}
		  \begin{tabular}[t]{c}
        % (Idea from reviewer comment in email 2018-06-08.)
				\includegraphics[width=.9\textwidth]{figures/scatter_dp_mc.png}
		\end{tabular}
  \end{center}
  \caption{{\bf Correspondance between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. We subsequently plotted the obtained $p$~values against each other. Again, the dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$.}
  \label{fig:pscatter}
{\em FIXME: Rename x/y-labels to GeneSetXX $p$~value}
\end{figure}

@@

LK was supported by a grant from Swedish Foundation for Strategic Research (BD15-0043).

>>>>

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*lukas.kall@scilifelab.se
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often referred to as pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks. While the terminology varies in the literature, we will here refer to such methods as Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations. Here we instead have designed one dynamic programming algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We also implemented a random sampling method, carrying out the same task. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.

@@

A limiting factor of ORA is the quality of the databases defining pathways. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the terminology varies in the literature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).

@@

Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}. We argue that such test is much easier implemented as tests of the query set, i.e. is the query set a randomly selected set of genes or is there an overrepresentation of well-connected genes in the query set?

@@

In network-based gene set analysis one scores a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ (i.e. $\mathcal{Q} \subseteq \mathcal{G}$), and how they relate to a pathway, $\mathcal{P}=\{p_1 \ldots p_P\}$. In NEA the pathway maps to the genome through a network, ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise. The pathways are scored by summing up all connections, i.e. the pathway score can be expressed as,

@@

We wanted to evaluate our score under the null hypothesis, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''. This formulation translates to calculating the distribution of scores that would occur if the $Q$ query genes are randomly selected from the genome, and see how frequently a score as extreme or more extreme than the current score would appear in that distribution. Hence, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtain a score, $s$, which we will give two methods for below. Given such a score distribution, we can express a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score, $s$, as,

@@

    for a in range(len(k)):
        for s in range(max_s,-1,-1):
            for c in range(Q,-1,-1):
                for b in range(k[a],0,-1): # Stop at b=1
                    if c-b>=0 and s-a*b>=0:
                        N[s,c] += comb(k[a], b) * N[s-a*b,c-b]
    return N[:,q]
\end{lstlisting}

There is a memory efficient implementation. We first note that

@@

To test the consistency of the methods, we also investigated the correspondence between the $p$~values reported by GeneSetDP and GeneSetMC. As can be seen in Figure~\ref{fig:pscatter}, the results are consistent as expected.

\begin{figure}[htb]
  \begin{center}
		  \begin{tabular}[t]{c}
        % (Idea from reviewer comment in email 2018-06-08.)
				\includegraphics[width=.9\textwidth]{figures/scatter_dp_mc.png}
		\end{tabular}
  \end{center}
  \caption{{\bf Correspondance between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway. We subsequently plotted the obtained $p$~values against each other. Again, the dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$.}
  \label{fig:pscatter}
% {\em FIXME: Rename x/y-labels to GeneSetXX $p$~value}
\end{figure}

@@

LK was supported by a grant from the Swedish Foundation for Strategic Research (BD15-0043).
